---
title: "Biostat 203B Homework 2"
author: "Ami Sheth"
subtitle: Due Feb 7 @ 11:59PM
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
```

```{r}
sessionInfo()
```

Use tidyverse (ggpot2, dplyr) to explore the [MIMIC-III](https://mimic.physionet.org) data introduced in [homework 1](https://ucla-biostat203b-2020winter.github.io/hw/hw1/hw1.html).

## Q1

Demographic information of the patients admitted into hospital is available in `ADMISSION.csv`. See <https://mimic.physionet.org/mimictables/admissions/> for details of each field in this file. Summarize following variables using appropriate graphs:   

- admission year  
- admission month  
- admission week day  
- admission hour  
- length of hospital stay  
- admission type  
- number of admissions per patient  
- admission location  
- insurance  
- language  
- religion  
- martial status  
- ethnicity  
- death 

Note it is possible that one patient (uniquely identified by the `SUBJECT_ID`) is admitted into hospital multiple times. When summarizing some demographic information, it makes sense to summarize based on only unique patients. 

**Solution:**

### Prerequistites 

Load tidyverse and lubridate:
```{r}
library(tidyverse)
library(lubridate)
```

### Import data
Let's first display the first 10 lines of the `ADMISSION.csv` file:
```{bash}
head /home/203bdata/mimic-iii/ADMISSIONS.csv
```

First try of parsing the data:
```{r}
admission %>% print(width = Inf)
admission <- read_csv("/home/203bdata/mimic-iii/ADMISSIONS.csv",
                      col_types = cols(ROW_ID = col_integer(),
                                       SUBJECT_ID = col_integer(),
                                       HADM_ID = col_integer(),
                                       HOSPITAL_EXPIRE_FLAG = col_logical(),
                                       HAS_CHARTEVENTS_DATA = col_logical()))
```
12 years diluted over 100 years span, number of hospital admissions is around 600 x 10 = 6000 admissions per year, can identify which years with lots of admissions (maybe why?), here not seeing much pattern, maybe investigation around 2130

86400 seconds = 1 year

## Summarizing the Data 

- admission year

```{r}
admission %>% mutate(adm_year = year(ADMITTIME)) %>% ggplot() + geom_bar(mapping = aes(x = adm_year)) + xlab("Year of Admission") + ggtitle("Barplot of Admission Year") + theme(plot.title = element_text(hjust = 0.5))
```

- admission month

```{r}
admission %>% mutate(adm_month = month(ADMITTIME, label = T)) %>% ggplot() + geom_bar(mapping = aes(x = adm_month)) + xlab("Month of Admission") + ggtitle("Barplot of Admission Month") + theme(plot.title = element_text(hjust = 0.5))
```

- admission week day

```{r}
admission %>% mutate(adm_day = day(ADMITTIME)) %>% ggplot() + geom_bar(mapping = aes(x = adm_day)) + xlab("Day of Admission") + ggtitle("Barplot of Admission Day") + theme(plot.title = element_text(hjust = 0.5))
```

- admission hour

```{r}
admission %>% mutate(adm_hour = hour(ADMITTIME)) %>% ggplot() + geom_bar(mapping = aes(x = adm_hour)) + xlab("Hour of Admission") + ggtitle("Barplot of Admission Hour") + theme(plot.title = element_text(hjust = 0.5))
```

- admission minute

```{r}
admission %>% mutate(adm_min = minute(ADMITTIME)) %>% ggplot() + geom_freqpoly(binwidth = 2, mapping = aes(x = adm_min)) + xlab("Minute of Admission") + ggtitle("Plot of Admission Minute") + theme(plot.title = element_text(hjust = 0.5))
```

- length of hospital stay 

```{r}
admission %>% mutate(stay_length = as.numeric(as.duration(DISCHTIME-ADMITTIME)/ 86400)) %>% ggplot() + geom_density(mapping = aes(x = stay_length)) + xlab("Length of Stay (days)") + ggtitle("Plot of Length of Hospital Stay") + theme(plot.title = element_text(hjust = 0.5))
```

- admission type  

```{r}
admission %>% ggplot() + geom_bar(mapping = aes(x = ADMISSION_TYPE)) + xlab("Admission Type") + ggtitle("Barplot of Admission Type") + theme(plot.title = element_text(hjust = 0.5))
```

- number of admissions per patient  

```{r}
admission %>% count(SUBJECT_ID) %>% ggplot() + geom_bar(mapping = aes(x = n)) + xlab("Number of Admissions") + ggtitle("Frequency of Admissions per Patient") + theme(plot.title = element_text(hjust = 0.5))
```

- admission location  

```{r}
admission %>% ggplot() + geom_bar(mapping = aes(x = ADMISSION_LOCATION)) + coord_flip() + xlab("Admission Location") + ggtitle("Barplot of Admission Location")
```

- insurance  

```{r}
admission %>% distinct(SUBJECT_ID, .keep_all = TRUE) %>% ggplot() + geom_bar(mapping = aes(x = INSURANCE)) + ggtitle("Plot of Patient Insurance") + theme(plot.title = element_text(hjust = 0.5))
```

- language  

```{r}
admission %>% distinct(SUBJECT_ID, .keep_all = TRUE) %>% ggplot() + geom_bar(mapping = aes(x = LANGUAGE)) + ggtitle("Plot of Patient Language") + theme(plot.title = element_text(hjust = 0.5), axis.text.x = element_text(angle = 90, hjust = 1))
```

- religion  

```{r}
admission %>% distinct(SUBJECT_ID, .keep_all = TRUE) %>% ggplot() + geom_bar(mapping = aes(x = RELIGION)) + ggtitle("Plot of Patient Religion") + theme(plot.title = element_text(hjust = 0.5), axis.text.x = element_text(angle = 45, hjust = 1))
```

- marital status 

```{r}
admission %>% distinct(SUBJECT_ID, .keep_all = TRUE) %>% ggplot() + geom_bar(mapping = aes(x = MARITAL_STATUS)) + ggtitle("Plot of Marital Status") + xlab("Marital Status") + theme(plot.title = element_text(hjust = 0.5), axis.text.x = element_text(angle = 45, hjust = 1))
```

- ethnicity  

```{r}
admission %>% distinct(SUBJECT_ID, .keep_all = TRUE) %>% ggplot() + geom_bar(mapping = aes(x = ETHNICITY)) + ggtitle("Plot of Patient Ethnicity") + theme(plot.title = element_text(hjust = 0.5), axis.text.x = element_text(angle = 45, hjust = 1))
```

- death 

```{r}
admission %>% count(DEATHTIME)
```


## Q2

Link the data in `ADMISSION.csv` and `PATIENTS.csv` (<https://mimic.physionet.org/mimictables/patients/>) and summarize following variables using appropriate graphs:  

- gender  
- age at admission 

## Q3

`ICUSTAYS.csv` (<https://mimic.physionet.org/mimictables/icustays/>) contains data about Intensive Care Units (ICU) stays. Summarize following variables using appropriate graphs:  

- length of ICU stay  
- first ICU unit  
- gender  
- age  

## Q4 

`CHARTEVENTS.csv` (<https://mimic.physionet.org/mimictables/chartevents/>) contains all the charted data available for a patient. During their ICU stay, the primary repository of a patient’s information is their electronic chart. The `ITEMID` variable indicates a single measurement type in the database. The `VALUE` variable is the value measured for `ITEMID`. 

`D_ITEMS.csv` (<https://mimic.physionet.org/mimictables/d_items/>) is the dictionary for the `ITEMID` in `CHARTEVENTS.csv`. Find potential values of `ITEMID` that correspond to systolic blood pressure, i.e., `LABEL` contains the string `systolic`. 

Compile a tibble that contains the first ICU stay of unique patients, with the patient's demographic information, the first systolic blood pressure measurement during ICU stay, and whether the patient died within 30 days of hospitcal admission.